---
title: "Investigate Red Wine Quality by Felix Lou"
author: "Felix Lou"
output: 
  html_document: 
    toc: true
    toc_depth: 3
    toc_float: true
editor_options: 
  chunk_output_type: console
---

========================================================

```{r echo=FALSE, message=FALSE, warning=FALSE, packages}
# Load all of the packages

library(ggplot2)
library(dplyr)
library(tidyr)
library(GGally)
library(scales)
library(memisc)
library(reshape2)
library(gridExtra)
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
df <- read.csv('wineQualityReds.csv')
```

# About
>In this project, I am going to explore a dataset in which contains information 
on red wine quality and chemical properties associated with them. With the help 
of the statistical program R, I am going to first conduct preliminary 
investigation to see if there are any relationships among variables, and further 
illustrate them with plots. The dataset is available for download [here](
https://www.google.com/url?q=https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityReds.csv&sa=D&ust=1516870686057000&usg=AFQjCNEc_chY2bDZIHK4qSD7WeKygHIAGQ), 
and its documentation is available [here](https://s3.amazonaws.com/udacity-hosted-downloads/ud651/wineQualityInfo.txt).

## Preliminary Investigation
>Let's run some basic functions to have a glance of the dataset. 

```{r echo=TRUE, Preliminary_Investigation}
# Check the general structure of the dataset
str(df)

# Double check to see if there are any dupicates
anyDuplicated(df)

# A glimpse on the statistical summaries on each variable
summary(df)
```

>Some findings at first glance:

- There are altogether 1599 observations of 13 numeric variables.
- `X` seems to be the ID of each wine
- Quality-wise, it seems most wines converge towards the middle on a 0-10 score scale, 
with a max of 8, min of 3, median of 6, and mean of 5.636. `quality` here is 
supposed to be a categorical vairable
- According to the documentation, some variables, like the acids, may be 
correlated with each other. This implies there may be multicollinearity
- Apart from `quality` and `.sulfur.dioxide`, all other variables are continuous

```{r echo=FALSE, results='hide', Update_Variable_Type}
# Change X to a factor (Nominal) variable
df$X <- factor(df$X)

# Sanity check
str(df$X)

# Change quality to a factor & ordinal variable
df$quality <- factor(df$quality, ordered = T)

# Sanity check
str(df$quality)
```

## A Glimpse of the Distribution
>Let's make a set of histograms for all the variables to have a general idea on 
the distribution first.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10, Univariate_Plots}
# Melt the dataframe and map 'value' as the x variable
ggplot(data = melt(df), 
       aes(x = value)) + 
  geom_histogram(color = 'black', 
                 fill = '#099DD9') + 
  facet_wrap(~variable, scales = 'free')

# Histogram on quality
qplot(x = df$quality, 
      color = I('black'), 
      fill = I('#099DD9'), 
      xlab = 'Quality')
```

# Univariate Analysis
## Quality of Wine
>Looking at the histogram of wine quality, we can see that most of the wines 
receive about average ratings which are based on a discrete range, and there are 
not really any extreme cases (Outliers). Although not obvious, there is some 
level of normal distribution.

>Based on this distribution, let's instantiate another variable on the basis of 
ratings. Three categories, 'good', 'average', and 'bad' will represent wines 
that receive rating of 7 or above, 5 or 6, and below 5 respectively.

```{r, echo=FALSE, message=FALSE}
# Instantiate categorical variable 'rating'
df$rating <- ifelse(df$quality < 5, 'bad', 
                    ifelse(df$quality < 7, 'average', 'good'))
df$rating <- ordered(df$rating,
                     levels = c('bad', 'average', 'good'))

# Sanity check
table(df$rating)
```

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Plot the rating
qplot(x = df$rating, 
      color = I('black'), 
      fill = I('#099DD9'), 
      xlab = 'Rating')
```

## Acidity of Wine
>Acids are major wine constituents and contribute greatly to its taste. 
According to the documentation and by researching 
[online](http://waterhouse.ucdavis.edu/whats-in-wine/fixed-acidity), 
`fixed.acidity` and `volatile.acidity` are two different types of acids 
(tartaric and acetic); so let's instantiate another variable that stands for 
overall acidity of wine.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Overall acidity variable
df$overall.acidity <- df$fixed.acidity + df$volatile.acidity + df$citric.acid

# Plot overall acidity
ggplot(aes(x = overall.acidity), 
       data = df) + 
  geom_histogram(color = 'black', 
                 fill = '#099DD9')
```


## High Level Comparison on Good and Bad
>Although by looking at the above histograms, we can tell the general 
properties that wines have, we cannot really tell what separates the good ones 
from the bad ones. 

>We need to perform a comparison in order to tell the difference. Let's first 
compare the overall acidity between the good ones and bad ones.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Comparison on overall acidity
ggplot(data = subset(df, rating != 'average'), 
       aes(x = overall.acidity, 
           fill = rating)) + 
  geom_density(alpha = 0.6)
```

>I believe there are reasons why wine experts divide acidity into two groups; so 
let's dig deeper by plotting the acids individually along with other variables.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Comparison on low level along with other variables
ggplot(data = melt(subset(df, rating != 'average')), 
       aes(x = value, fill = rating)) + 
  geom_density(alpha = 0.6) + 
  facet_wrap(~variable, scales = 'free')
```

>Here, we plot the probability density functions of the good ones and the bad 
ones. Generally speaking, we want to focus on the area where the two groups do 
not overlap becasue that can serve as a reference of what separates the good 
ones from the bad ones. In other words, variables in which the two groups have 
obvious difference in distribution could be indicators of wine quality, or more 
or less help us predict wine quality.

>Looking at the above plots, we can see there are relatively obvious differences
in `volatile.acidity`, `citric.acid`, `pH`, `sulphates`, and `alcohol`. And if 
we look at the plots individually, we can see that the ranges of spikes are 
especially wide among `citric.acid` and `alcohol` (On its own scale). This 
suggests that the two variables might serve as indicators of wine quality.

## Outliers
>As we saw from the plots above, some variables' distribution is skewed. This 
suggests that there might be outliers in the dataset. Let's validate this with 
the help of boxplot.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Boxplot to spot outliers
ggplot(data = melt(df), 
       aes(x = 'Wine', y = value)) + 
  geom_boxplot(color = 'black', 
                 fill = '#099DD9') + 
  facet_wrap(~variable, scales = 'free')
```

>Now we have a much clearer picture on outliers. 

# Bivariate Analysis

>As mentioned in the beginning, there may be multicollinearity among variables; 
so let's get a quick snapshot of the correlations among them.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=10}
# Correlation table among variables
ggcorr(df)
```

>Indeed, multicollinearities exist in this dataset. Basically, the darker/ 
sharper the grid, the stronger the correlations associated with the pair of 
variables. One interesting finding is that there seems to be a negative 
correlation between `citric.acid` and `volatile.acidity`, and `volatile.acidity` 
seems to have a slightly positive relationship with `pH`. This is actually 
quite counterintuitive. 

>From the density plots, we can see some variables may have an impact on wine 
quality. Let's see if we can capture the trend/ tendency with the help of 
boxplot.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Boxplot on rating
ggplot(data = melt(df), 
       aes(x = rating, 
           y = value)) + 
  geom_jitter(alpha = 0.2) + 
  geom_boxplot(alpha = 0.4, color = 'blue') + 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
  xlab('Rating') + 
  facet_wrap(~variable, scales = 'free')

# Boxplot on quality
ggplot(data = melt(df), 
       aes(x = quality, 
           y = value)) + 
  geom_jitter(alpha = 0.2) + 
  geom_boxplot(alpha = 0.4, color = 'blue') + 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
  xlab('Quality') + 
  facet_wrap(~variable, scales = 'free')
```

>Now the relationships between certain variables and wine quality are much 
clearer. In general, the steeper the boxes are positioned against each other, 
the greater the impact of that specific variable on wine quality. Here, 
`fixed.acidity`, `volatile.acidity`, `pH`, `citric.acid`, `sulphates`, and 
`alcohol` all seem to have impact on wine quality. This is especially true for `alcohol`, `sulphates`, `volatile.acidity`, and `citric.acid`. Let's double 
check by calculating the correlation coefficient of each variable against 
`quality`.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Create a boolean filter for numeric columns
df_num_columns <- sapply(df, is.numeric)

# Correlation of variables (Numeric) against quality
cor(df[ , df_num_columns], as.numeric(df$quality))
```

>To summarize:

- Good wines usually have relatively high fixed acidity, including citric acid
- Good wines usually have relatively high alcohol level
- Good wines usually have relatively high sulphates level
- Good wines usually have relatively low volatile acidity

>As mentioned previously, although good wines generally have 
relatively high acidity, they also in general possess lower level of 
`volatile.acidity`. Looking into the documentation, the amount of acetic acid 
in wine, which at too high of levels can lead to an unpleasant, vinegar taste. 
It seems `volatile.acidity` is negatively correlated with other acids. 

>Let's plot `volatile.acidity` against `citric.acid` and `fixed.acidity`.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Scatter plot on volatile.acidity against citric.acid
ggplot(data = df, 
       aes(x = volatile.acidity, 
           y = citric.acid)) + 
  geom_point(alpha = 0.5, 
             size = 1, 
             color = '#099DD9', 
             position = position_jitter(height = 0)) + 
  geom_smooth(method = 'lm', 
              color = 'red', 
              size = 1, 
              se = FALSE)

# Correlation coefficient
with(df, cor.test(df$volatile.acidity, df$citric.acid))

# Scatter plot on volatile.acidity against fixed.acidity
ggplot(data = df, 
       aes(x = volatile.acidity, 
           y = fixed.acidity)) + 
  geom_point(alpha = 0.5, 
             size = 1, 
             color = '#099DD9', 
             position = position_jitter(height = 0)) + 
  geom_smooth(method = 'lm', 
              color = 'red', 
              size = 1, 
              se = FALSE)

# Correlation coefficient
with(df, cor.test(df$volatile.acidity, df$fixed.acidity))
```

>`volatile.acidity` is indeed negatively correlated with other acids. 
This is especially true for `citric.acid`, given the correlation coefficient is 
-0.552.

>According to the documentation, `total.sulfur.dioxide` includes 
`free.sulfur.dioxide`. This implies that there should be a correlation between 
the two variables.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Scatter plot on `total.sulfur.dioxide` against `free.sulfur.dioxide`
ggplot(data = df, 
       aes(x = free.sulfur.dioxide, 
           y = total.sulfur.dioxide)) + 
  geom_point(alpha = 0.5, 
             size = 1, 
             color = '#099DD9', 
             position = position_jitter(height = 0)) + 
  geom_smooth(method = 'lm', 
              color = 'red', 
              size = 1, 
              se = FALSE)

# Correlation coefficient
with(df, cor.test(df$free.sulfur.dioxide, df$total.sulfur.dioxide))
```

>As expected, there is such a strong positive correlation of 0.668 between the 
two variables.

>According to the documentation, `density` is a variable that depends on 
alcohol level and sugar content. This suggests that there may be a correlation 
between `density` against `residual.sugar` and `alcohol`.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Scatter plot on `residual.sugar` against `density`
ggplot(data = df, 
       aes(x = residual.sugar, 
           y = density)) + 
  geom_point(alpha = 0.5, 
             size = 1, 
             color = '#099DD9', 
             position = position_jitter(height = 0)) + 
  geom_smooth(method = 'lm', 
              color = 'red', 
              size = 1, 
              se = FALSE)

# Correlation coefficient
with(df, cor.test(df$residual.sugar, df$density))

# Scatter plot on `alcohol` against `density`
ggplot(data = df, 
       aes(x = alcohol, 
           y = density)) + 
  geom_point(alpha = 0.5, 
             size = 1, 
             color = '#099DD9', 
             position = position_jitter(height = 0)) + 
  geom_smooth(method = 'lm', 
              color = 'red', 
              size = 1, 
              se = FALSE)

# Correlation coefficient
with(df, cor.test(df$alcohol, df$density))
```

>The results are as what we expected. `residual.sugar` is positively correlated 
with `density`, and `alcohol` is negatively correlated with `density`.

>pH should be hugely affected by wines acidity as `pH` is essentially a measure 
of acidity; so let's plot `overall.acidity` against `pH`.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Scatter plot on `overall.acidity` against `pH`
ggplot(data = df, 
       aes(x = overall.acidity, 
           y = pH)) + 
  geom_point(alpha = 0.5, 
             size = 1, 
             color = '#099DD9', 
             position = position_jitter(height = 0)) + 
  geom_smooth(method = 'lm', 
              color = 'red', 
              size = 1, 
              se = FALSE)

# Correlation coefficient
with(df, cor.test(df$overall.acidity, df$pH))
```

>The results are as expected. `overall.acidity` is negatively correlated with 
`pH`.

# Multivariate Analysis

>The last plot seems to be the most linear compared to others. Sush a strong 
coorelation suggests that we might use a linear model to predict `pH` based on 
`overall.acidity`.

```{r echo=FALSE, message=FALSE, fig.width=14, fig.height=10}
# Build and fit the linear model for `pH` and `overall.acidity`
model <- lm(I(pH) ~ I(overall.acidity), data = df)

# Prediction
df$pH.estimate <- predict(model, df)

# Measure the difference between true values and estimates
df$pH.residual <- (df$pH.estimate - df$pH) / df$pH

# Boxplot on residuals
ggplot(data = df, 
       aes(x = quality, 
           y = pH.residual)) +
  geom_jitter(alpha = 0.5) + 
  geom_boxplot(alpha = 0.5, color = 'blue') + 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
  xlab('Quality')
```

>According to the boxplot, the residuals are generally consistent across 
qualities, except the ones with the lowest quality. The ones with quality 3 has 
a median well below 0. Supposedly, acidity itself should be the biggest factor 
that alters `pH`. Nonetheless, this does not really apply to this group of wine. 
Chances are there might be omitted variables.

>Previously, we spotted numbers of correlations among variables. Now let's 
visualize them on rating and quality basis, and see if the pattern is 
consistant across groups.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=10, Multivariate_Plots}

ggplot(data = df, 
       aes(x = volatile.acidity, 
           y = citric.acid, 
           color = quality)) +
      geom_point(alpha = 0.8, 
                 size = 1, 
                 position = position_jitter(height = 0)) +
      geom_smooth(method = "lm", 
                  size = 1, 
                  se = FALSE)  +
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Quality')) + 
  facet_wrap(~rating) + 
  theme_dark()

ggplot(data = df, 
       aes(x = free.sulfur.dioxide, 
           y = total.sulfur.dioxide, 
           color = quality)) +
      geom_point(alpha = 0.8, 
                 size = 1, 
                 position = position_jitter(height = 0)) +
      geom_smooth(method = "lm", 
                  size = 1, 
                  se = FALSE)  +
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Quality')) + 
  facet_wrap(~rating) + 
  scale_y_continuous(lim = c(0, quantile(df$total.sulfur.dioxide, 0.99))) + 
  theme_dark()

ggplot(data = df, 
       aes(x = alcohol, 
           y = density, 
           color = quality)) +
      geom_point(alpha = 0.8, 
                 size = 1, 
                 position = position_jitter(height = 0)) +
      geom_smooth(method = "lm", 
                  size = 1, 
                  se = FALSE)  +
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Quality')) + 
  facet_wrap(~rating) + 
  theme_dark()

ggplot(data = df, 
       aes(x = overall.acidity, 
           y = pH, 
           color = quality)) +
      geom_point(alpha = 0.8, 
                 size = 1, 
                 position = position_jitter(height = 0)) +
      geom_smooth(method = "lm", 
                  size = 1, 
                  se = FALSE)  +
  scale_color_brewer(type = 'seq', 
                     guide = guide_legend(title = 'Quality')) + 
  facet_wrap(~rating) + 
  theme_dark()
```

>The patterns are consistent across groups.

>Previously, we found that the ranges of spikes are especially wide among 
`citric.acid` and `alcohol` (On its own scale) when doing univariate analysis. 
 We also found that there are relatively strong correlations between `quality`, 
`volatile.acidity`, and `sulphates`. Let's see how the good ones and bad ones 
are distributed in a scatter plot.

```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=10}
ggplot(data = subset(df, rating != 'average'), 
       aes(x = citric.acid, 
           y = alcohol, 
           color = rating)) + 
  geom_point(position = position_jitter())

ggplot(data = subset(df, rating != 'average'), 
       aes(x = volatile.acidity, 
           y = alcohol, 
           color = rating)) + 
  geom_point(position = position_jitter())

ggplot(data = subset(df, rating != 'average'), 
       aes(x = sulphates, 
           y = alcohol, 
           color = rating)) + 
  geom_point(position = position_jitter())
```

>By plotting them in scatter plots, our ideas are further validated. Although 
the plots do not have a crystal clear generalization, they provide us with a 
pretty clear picture on what separate good wines from bad wines.

------

# Final Plots and Summary

### Plot One - Probability Density Function on Citric Acid and Alcohol
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=10, Plot_One}
p1a <- ggplot(aes(x = citric.acid, 
                  fill = rating), 
              data = subset(df, rating != 'average')) + 
  geom_density(alpha = 0.6) + 
  xlab('Citric Acid (g / dm^3)') + 
  ylab('Probability Density Function') + 
  ggtitle('Probability Density Function of Citric Acid Level')

p1b <- ggplot(aes(x = alcohol, 
                  fill = rating), 
              data = subset(df, rating != 'average')) + 
  geom_density(alpha = 0.6) + 
  xlab('Alcohol (% by volume)') + 
  ylab('Probability Density Function') + 
  ggtitle('Probability Density Function of Alcohol Level')

grid.arrange(p1a, p1b, ncol = 1)
```

>This first pair of plots shows how wine groups are distributed based on the 
variables. The variable is a fairly good indicator when the distributions do 
not overlap and start to get farther from each other.

### Plot Two - Volatile Acidity & Sulphates Level against Quality
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=10, Plot_Two}
p2a <- ggplot(data = df, 
              aes(x = quality, 
                  y = volatile.acidity)) + 
  geom_jitter(alpha = 0.2) + 
  geom_boxplot(alpha = 0.4, color = 'blue') + 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
  scale_y_continuous(lim = c(0, quantile(df$volatile.acidity, 0.99))) + 
  xlab('Wine Quality') + 
  ylab('Volatile acidity (acetic acid - g / dm^3)') + 
  ggtitle('Volatility Acidity Level among Different Wine Qualities')

p2b <- ggplot(data = df, 
              aes(x = quality, 
                  y = sulphates)) + 
  geom_jitter(alpha = 0.2) + 
  geom_boxplot(alpha = 0.4, color = 'blue') + 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4) + 
  scale_y_continuous(lim = c(0, quantile(df$sulphates, 0.95))) + 
  xlab('Wine Quality') + 
  ylab('Sulphates (Potassium Sulphate - g / dm3)') + 
  ggtitle('Sulphates Level among Different Wine Qualities')

grid.arrange(p2a, p2b, ncol = 1)
```

>This second pair of plots shows how a drop in `valitility.acidity` and a rise 
in `sulphates` can enhance wines' quality. Although they cannot tell the whole 
story, the plots are good enough for us to say that these two variables should 
not be neglected.


### Plot Three - Good & Bad Wines Generalization
```{r echo=FALSE, message=FALSE, warning=FALSE, fig.width=14, fig.height=10, Plot_Three}
p3a <- ggplot(data = subset(df, rating != 'average'), 
       aes(x = citric.acid, 
           y = alcohol, 
           color = rating)) + 
  geom_point(position = position_jitter()) + 
  xlab('Citric acid (g / dm^3)') + 
  ylab('Alcohol (% by volume)') + 
  ggtitle('Citric Acid & Alcohol on Wine Ratings')

p3b <- ggplot(data = subset(df, rating != 'average'), 
       aes(x = volatile.acidity, 
           y = alcohol, 
           color = rating)) + 
  geom_point(position = position_jitter()) + 
  xlab('Volatile acidity (acetic acid - g / dm^3)') + 
  ylab('Alcohol (% by volume)') + 
  ggtitle('Volatility Acidity & Alcohol on Wine Ratings')

grid.arrange(p3a, p3b, ncol = 1)
```

>This third pair of plots further demonstrates the idea of the first pair of 
plots. Here, we can clearly see that where certain wines are clustered. We can 
be even more confident to say that these variables are some of the key factors 
to consider when evaluating wine.

------

# Reflection

>In this EDA, I was able to identity some of the key factors that have impact on 
wine quality. Although wine quality is arguably subjective, the results we got 
from this analysis are reasonable. At least we can say that these vairables that 
we investigated do play a role on wine quality according to conventional/ 
industry standard. Generally speaking, acidity, sulphates level, and alcohol 
level are the ones that could alter wine quality most. There could be more 
probabilistic work if we want to run hypothesis testing on certain statistical 
summaries like the difference in alcohol means among groups.
>One thing that is worth mentioning is multicollinearity is found in the dataset. 
If we want to build a linear model, ideally speaking, we want the predictor 
variables to be related to the response variable, but not to be related to one 
another. Multicollinearity occurs when predictor variables are related to one 
another. Expected relationships may not hold under the presence of 
Multicollinearity. Worse, any hypothesis testing becomes unreliable. As we saw in 
the correlation table, quite a few varibles are correlated with one another. 
If we are to build a multi-linear regression model, we would have to deal with 
this issue. One common way to working with correlated predictor variables in a multi-linear regression model, is simply to remove one of the variables that is 
most related to the other variables. Choosing an predictor variable that is not 
as important is a common choice.